Symmetric Circuits for Rank Gates

### 2018-02-05 sections/transparency-and-unique-extensions.tex
February 5 2018

% consider only symmetric circuits with symmetric gates, which are transparent,
% and as such include no such condition.

@@

% \begin{definition}
%   We say a circuit $C_n$ is que labels it has unique extensions. We finally
%   discuss th\emph{rigid} if it has injective labels and for every pair of
%   gates $(g, g')$ in $C_n$ such that $H_g = H_{g'}$ and $g \equiv g'$.
% \end{definition}
\subsection{The Usefulness of Transparency}
In the this subsection we show that a number of important circuit properties are
polynomial-time decidable for transparent circuits (e.g. symmetry, the
syntactic-equivalence relation, etc.). We also provide polynomial-time
algorithms for computing the action of a permutation on a symmetric circuit, as
well as for computing the orbits and supports of gates, and elements of
universes of gates, in a circuit.

@@

\begin{lem}
  There is an algorithm that takes as input a transparent circuit $C$ and
  outputs the syntactic-equivalence relation on the gates of $C$. The algorithm
  runs in time polynomial in the size of $C$.
  \label{lem:unique-labels-syntactic-equiv}
\end{lem}
\begin{proof}
  Let $C := \langle G, \Omega, \Sigma, \Lambda, L \rangle$ be a transparent
  $(\mathbb{B}, \rho)$-circuit of order $n$. We define a relation $\sim$ by
  induction on height (i.e. distance from an input gate). This relation will be
  the syntactic-equivalence relation on $G$. It is easy to check the conditions
  for syntactic-equivalence for gates of height 0 as all such gates are input
  gates and so are either constant gates or relational gates. Suppose $g$ and
  $h$ are gates of height greater than zero. We only need to check conditions
  (i), (iv) and (v) of syntactic-equivalence. It is easy to check (i) and (v).
  If either condition is false then $g$ and $h$ are not syntacticly equivalent.
  So suppose $g$ and $h$ are two internal gates such that $\Sigma (g) =
  \Sigma(h)$ and suppose that if they are both output gates then $\Omega^{-1}(g)
  = \Omega^{-1}(h)$. It remains to check if $L(g)/{\sim}$ and $L(h)/{\sim}$ are
  isomorphism equivalent. First check if $H_h /{\sim} = H_g /{\sim}$. If not,
  $g$ and $h$ are not syntactically equivalent. So suppose $H_h /{\sim} = H_g
  /{\sim}$. If $g$ and $h$ are symmetric gates then $H_h /{\sim} = H_g/{\sim}$
  if, and only if, $L(g)/{\sim}$ and $L(h)/{\sim}$ are isomorphism equivalent,
  so let $g \sim h$. If $g$ and $h$ are not symmetric gates then, since $C$ is
  transparent, $g$ and $h$ have unique labels. We thus have that $L(g)/{\sim}$
  and $L(h)/{\sim}$ are isomorphism equivalent if, and only if, $L(g)^{-1}L(h)$
  acts like an isomorphism. This can be checked by first checking that the
  function induces a sorted permutation on the universe of $g$ (which can be
  checked efficiently). This is sufficient as $L(g)(L(g)^{-1}L(h)) = L(h)$.

  It is easy to see that, given a reasonable binary encoding of a circuit, the
  above procedure can be implemented so as to run in polynomial time, and that
  the relation $\sim$ is the syntactic-equivalence relation on the gates of $C$.

  % Let $C := \langle G, \Omega, \Sigma, \Lambda, L \rangle$ be a transparent
  % $(\mathbb{B}, \rho)$-circuit of order $n$. We define a relation $\sim$ by
  % induction on height (i.e. distance from an input gate). This relation will be
  % the syntactic-equivalence relation on $G$. It is easy to check the conditions
  % for syntactic-equivalence for gates of height 0 as all such gates are input
  % gates and so are either constant gates or relational gates. Suppose $g$ and
  % $h$ are gates of height greater than zero. We only need to check conditions
  % (i), (iv) and (v) of syntactic-equivalence. It is easy to check (i) and (v).
  % If either condition is false then $g$ and $h$ are not syntacticly equivalent.
  % So suppose $g$ and $h$ are two internal gates such that $\Sigma (g) =
  % \Sigma(h)$ and suppose that if they are both output gates then $\Omega^{-1}(g)
  % = \Omega^{-1}(h)$. It remains to check if $L(g)/{\sim}$ and $L(h)/{\sim}$ are
  % isomorphism equivalent. First check if $H_h /{\sim} = H_g /{\sim}$. If not,
  % $g$ and $h$ are not syntactically equivalent. So suppose $H_h /{\sim} = H_g
  % /{\sim}$. If $g$ and $h$ are symmetric gates then $H_h /{\sim} = H_g/{\sim}$
  % if, and only if, $L(g)/{\sim}$ and $L(h)/{\sim}$ are isomorphism equivalent,
  % so let $g \sim h$. If $g$ and $h$ are not symmetric gates then, since $C$ is
  % transparent, $g$ and $h$ have unique labels. We thus have that $L(g)/{\sim}$
  % and $L(h)/{\sim}$ are isomorphism equivalent if, and only if, $L(g)^{-1}L(h)$
  % acts like an isomorphism. This can be checked by first checking that the
  % function induces a sorted permutation on the universe of $g$ (which can be
  % checked efficiently). This is sufficient as $L(g)(L(g)^{-1}L(h)) = L(h)$.

  % It is easy to see that, given a reasonable binary encoding of a circuit, the
  % above procedure can be implemented so as to run in polynomial time, and that
  % the relation $\sim$ is the syntactic-equivalence relation on the gates of $C$.

>>>>

% consider only symmetric circuits with symmetric gates, which are transparent,
% and as such include no such condition

@@

% \begin{definition}
%   We say a circuit $C_n$ is que labels it has unique extensions. We finally
%   discuss th\emph{rigid} if it has injective labels and for every pair of
%   gates $(g, g')$ in $C_n$ such that $H_g = H_{g'}$ and $g \equiv g'$.
% \end{definition}

\subsection{The Usefulness of Transparency}
In the this subsection we show that a number of important circuit properties are
polynomial-time decidable for transparent circuits (e.g. symmetry, the
syntactic-equivalence relation, etc.). We also provide polynomial-time
algorithms for computing the action of a permutation on a symmetric circuit, as
well as for computing the orbits and supports of gates, and elements of
universes of gates, in a circuit.

@@

\begin{lem}
  There is an algorithm that takes as input a transparent circuit $C$ and
  outputs the syntactic-equivalence relation on the gates of $C$. The algorithm
  runs in time polynomial in the size of $C$.
  \label{lem:unique-labels-syntactic-equiv}
\end{lem}
\begin{proof}
  Let $C := \langle G, \Omega, \Sigma, \Lambda, L \rangle$ be a transparent
  $(\mathbb{B}, \rho)$-circuit of order $n$. We define the syntactic-equivalence
  relation on $G$ by induction on the height of a gate We note that this
  definition can be implemented by a dynamic program.

  Suppose $g$ and $h$ are gates in $C$ such that $\Sigma (g) = \Sigma(h)$ and
  either both $g$ and $h$ are output gates and $\Omega^{-1}(g) = \Omega^{-1}(h)$
  or neither are output gates. Let $g$ and $h$ be input gates, then set $g
  \equiv h$ if, and only if, both $g$ and $h$ are constant gates or both are
  relational gates and $\Lambda(g) = \Lambda (h)$. Let $g$ and $h$ be internal
  gates and suppose we have defined the syntactic-equivalence relation for all
  gates of height less than the height of either $g$ or $h$. Then $g \equiv h$
  if, and only if, $L(g)$ and $L(h)$ are isomorphism equivalent up to $\equiv$.

  % Let $C := \langle G, \Omega, \Sigma, \Lambda, L \rangle$ be a transparent
  % $(\mathbb{B}, \rho)$-circuit of order $n$. We define the
  % syntactic-equivalence
  % relation on $G$ by induction on the height of a gate. We note that this
  % definition can be implemented as a recursive algorithm.

  % Suppose $g$ and $h$ are gates in $C$, and check if $\Sigma (g) = \Sigma(h)$
  % and either both $g$ and $h$ are output gates and $\Omega^{-1}(g) =
  % \Omega^{-1}(h)$ or neither are output gates. If this check fails, return
  % that
  % $g \not\equiv h$. Otherwise, check if $g$ and $h$ are input gates and, if
  % so,
  % set $g \equiv h$ if, and only if, both $g$ and $h$ are constant gates or
  % both
  % are relational gates and $\Lambda(g) = \Lambda (h)$. Check if $g$ and $h$
  % are
  % internal gates and, if so, set $g \equiv h$ if, and only if, $L(g)$ and
  % $L(h)$
  % are isomorphism equivalent up to $\equiv$. ()

  The above algorithm can be implemented by a dynamic program with the number of
  iterations bounded by the number of pairs of gates in $C$, i.e. $\vert C
  \vert$ choose $2$. It remains to show that each iterative step can be
  implemented so as to run in time polynomial in the size of $C$. The only
  condition that is not obviously polynomial-time decidable is checking if
  $L(g)$ is isomorphism-equivalent to $L(h)$ up to $\equiv$. We sketch an
  algorithm that decides this condition in time polynomial in the size of the
  circuit.

  Let $g$ and $h$ be two internal gates in $C$ and suppose we have defined the
  syntactic-equivalence relation for all gates of height less than the height of
  either $g$ or $h$. We test if for each gate in $g' \in H_g$, there exists
  $h_{g'} \in H_h$ such that $g' \equiv h_{g'}$ and $\vert \{g'' \in H_g : g''
  \equiv g'\} \vert = \vert \{h'' \in H_h : h'' \equiv h'\}$. If this test
  fails, we set $g \not\equiv h$.

  We then check if both $g$ and $h$ are symmetric gates or if both are
  non-symmetric gates. Suppose both $g$ and $h$ are symmetric gates. We have
  that $L(g)$ is isomorphism-equivalent to $L(h)$ up to $\equiv$ if, and only
  if, there exists a bijection $f : H_g \rightarrow H_h$ such that $f(g') \equiv
  g'$, for all $g' \in H_g$ if, and only if, for each gate in $g' \in H_g$,
  there exists $h' \in H_h$ such that $g' \equiv h'$ and $\vert \{g'' \in H_g :
  g'' \equiv g'\} \vert = \vert \{h'' \in H_h : h'' \equiv h'\}$. Since we have
  already checked this condition we set $g \equiv h$.

  Suppose $g$ and $h$ are non-symmetric gates. Then, since $C$ is transparent,
  both $g$ and $h$ have unique labels. We thus have that for each $g' \in H_g$
  there is exactly one $h_{g'} \in H_h$ such that $g' \equiv h_{g'}$, and so we
  can define the bijection $f : H_g \rightarrow H_h$ by $f(g') := h_{g'}$ for
  all $g' \in H_g$. It can be shown that $L(g)$ is isomorphism-equivalent to
  $L(h)$ up to $\equiv$ if, and only if, $f \circ L(g)$ is
  isomorphism-equivalent to $L(h)$ if, and only if, $L(h)^{-1} f L(g)$ acts on
  $\ind (g)$ like an element of $\aut(g)$. We have from
  Lemma~\ref{lem:aut-partial} that $\sigma := L(h)^{-1} f L(g)$ acts like an
  element of $\aut(g)$ if, and only if, for all $\vec{a}_R, \vec{b}_{R'} \in \ind(g)$,
  $\sigma$ acts like a partial automorphism on $(\vec{a}_R, \vec{b}_{R'})$. We
  thus conduct the test by iterating through every pair $(\vec{a}_R,
  \vec{b}_{R'}) \in \ind(g)$, and set $g \equiv h$ if, and only if, $\sigma$
  acts like a partial automorphism on every such pair.

  % for all $\vec{a}_R \in \ind(g)$, $\sigma \vec{a}_R$ has type $R$, and for all
  % $\vec{b}_{R'}$, $i \in [\arty(R)]$, and $j \in [\arty(R')]$ we have that
  % $\vec{a}(i) = \vec{b}(j)$ if, and only if, $(\sigma \vec{a}_R)(i) = (\sigma
  % \vec{b}_R)(j)$. Since $\vert \ind(g) \vert \leq \vert C \vert$, we can check
  % if $\sigma$ acts like an element of $\aut(g)$ by iterating through $\ind(g)$
  % and checking each tuple satisfies the above conditions. We set $g \equiv h$
  % if, and only if, $\sigma$ acts like an element of $\aut(g)$.

  % This relation will be the syntactic-equivalence relation on $G$.

  % Let $g, h \in G$. Suppose $g$ and $h$ are input gates, then $g \sim h$ if,
  % and only if, $\Sigma (g) = \Sigma (h)$,
  % \item if $g$ and $h$ are relational gates then $\Lambda (g) = \Lambda (h)$,
  % \item either both or neither of $g$ and $h$ are output gates, and
  % \item if both $g$ and $h$ are output gates then $\Omega^{-1}(g) =
  %   \Omega^{-1}(h)$.
  % \end{itemize}
  % \setlength\itemsep{0mm} Suppose $g$ and $h$ are internal gates, and suppose
  % we have defined $\sim$ for all gates of height less than either $g$ or $h$.
  % Then $g \sim h$ if, and only if,
  % \begin{itemize}
  % \item $\Sigma (g) = \Sigma (h)$,
  % \item either both or neither of $g$ and $h$ are output gates,
  % \item if both $g$ and $h$ are output gates then $\Omega^{-1}(g) =
  %   \Omega^{-1}(h)$, and
  % \item $L (g)$ is isomorphism-equivalent to $L(h)$ up to $\sim$.
  % \end{itemize}

  % The above inductive definition can be implemented as a recursive algorithm
  % with the number of recursive steps bounded by the number of pairs of gates
  % in
  % $C$, i.e. $\vert C \vert$ choose $2$. It remains to show that each recursive
  % call can be implemented so as to run in time polynomial in the size of $C$.
  % It
  % is easy to see that we can check if a pair of input gates are
  % syntactically-equivalent in time polynomial in the size of $C$. The only
  % condition that is not obviously polynomial-time decidable is checking if
  % $L(g)$ is isomorphism-equivalent to $L(h)$ up to $\sim$. We sketch an
  % algorithm for deciding this condition for two internal gates.

  % Let $g$ and $h$ be two internal gates in $C$. We first check if both $g$ and
  % $h$ are symmetric gates or non-symmetric gates. Suppose both $g$ and $h$ are
  % symmetric gates. It can be shown that $L(g)$ is isomorphism-equivalent to
  % $L(h)$ up to $\sim$ if, and only if, there exists a bijection $f : H_g
  % \rightarrow H_h$ such that $f(g') \sim g'$, for all $g' \in H_g$ if, and
  % only
  % if, for each gate in $g' \in H_g$, there exists $h' \in H_h$ such that $g'
  % \sim h'$ and $\vert [g']_{\sim}\cap H_g \vert = \vert [h']_{\sim} \cap H_h$.
  % It is easy to see that this last condition can be checked in polynomial
  % time,
  % and so we can check if $g \sim h$ in polynomial-time.

  % Suppose $g$ and $h$ are non-symmetric gates. Then, since $C$ is transparent,
  % $g$ and $h$ have unique labels. First check if there exists for each $g' \in
  % H_g$ a gate $h_{g'} \in H_h$ such that $g' \sim h_{g'}$. If there exists
  % $g'$
  % such that no corresponding $h_{g'}$ exists, then we return that $g \not\sim
  % h$. Since $g$ and $h$ have unique labels for each $g' \in H_g$ there is at
  % most one $h_{g'} \in H_h$, and so we can define $f : H_g \rightarrow H_h$ by
  % $f(g') := h_{g'}$ for all $g' \in H_g$. We note that there exists $\sigma
  % \in
  % aut(g)$ such that $f(L(g)(\vec{a}_R)) \sim L(h)(\sigma \vec{a}_R)$ for all
  % $\vec{a}$, if and only if, $L(h)^{-1} f L(g)$ is compatible with every pair
  % $(h', h'') \in H^2_h$.

  % If this algorthm terminates and $f$ is constructed then we have that $L(g)$
  % is
  % isomorphism-equivalent to $L(h)$ up to $\sim$ and if this Suppose both $g$
  % and
  % $h$ are non-symmetric gates. Then both have unique labels, and so there is
  % exactly

  % Suppose $g$ is a symmetric gate. We then have that $L(g)$ is
  % isomorphism-equivalent to $L(h)$

  % It easy to see that for any pair of input gates we can check $g \sim h$ in
  % polynomial time. If $\Sigma (g) \neq \Sigma (h)$ then set $g \not\sim h$. If
  % exactly one of the two gates is an output gate then set $g \not\sim h$. If
  % both $g$ and $h$ are output gates and $\Omega^{-1}(g) \neq \Omega^{-1}(h)$,
  % then set $g \not\sim h$. If $g$ and $h$ are both relational gates and
  % $\Lambda
  % (g) \neq \Lambda (h)$ then set $g \not\sim h$.

  % We have thus far handled the base case.

  % If $g$ and $h$ are not both internal gates or both not input gates then set
  % $g
  % \not\sim h$. If $g$ and $h$ are both

  % is easy to check the conditions for syntactic-equivalence for gates of
  % height
  % 0 as all such gates are input gates and so are either constant gates or
  % relational gates. Suppose $g$ and $h$ are gates of height greater than zero.
  % We only need to check conditions (i), (iv) and (v) of syntactic-equivalence.
  % It is easy to check (i) and (v). If either condition is false then $g$ and
  % $h$
  % are not syntacticly equivalent. So suppose $g$ and $h$ are two internal
  % gates
  % such that $\Sigma (g) = \Sigma(h)$ and suppose that if they are both output
  % gates then $\Omega^{-1}(g) = \Omega^{-1}(h)$. It remains to check if
  % $L(g)/{\sim}$ and $L(h)/{\sim}$ are isomorphism equivalent. First check if
  % $H_h /{\sim} = H_g /{\sim}$. If not, $g$ and $h$ are not syntactically
  % equivalent. So suppose $H_h /{\sim} = H_g /{\sim}$. If $g$ and $h$ are
  % symmetric gates then $H_h /{\sim} = H_g/{\sim}$ if, and only if,
  % $L(g)/{\sim}$
  % and $L(h)/{\sim}$ are isomorphism equivalent, so let $g \sim h$. If $g$ and
  % $h$ are not symmetric gates then, since $C$ is transparent, $g$ and $h$ have
  % unique labels. We thus have that $L(g)/{\sim}$ and $L(h)/{\sim}$ are
  % isomorphism equivalent if, and only if, $L(g)^{-1}L(h)$ acts like an
  % isomorphism. This can be checked by first checking that the function induces
  % a
  % sorted permutation on the universe of $g$ (which can be checked
  % efficiently).
  % This is sufficient as $L(g)(L(g)^{-1}L(h)) = L(h)$.

  % It is easy to see that, given a reasonable binary encoding of a circuit, the
  % above procedure can be implemented so as to run in polynomial time, and that
  % the relation $\sim$ is the syntactic-equivalence relation on the gates of
  % $C$.

  % Let $C := \langle G, \Omega, \Sigma, \Lambda, L \rangle$ be a transparent
  % $(\mathbb{B}, \rho)$-circuit of order $n$. We define a relation $\sim$ by
  % induction on height (i.e. distance from an input gate). This relation will
  % be the syntactic-equivalence relation on $G$. It is easy to check the
  % conditions for syntactic-equivalence for gates of height 0 as all such gates
  % are input gates and so are either constant gates or relational gates.
  % Suppose $g$ and $h$ are gates of height greater than zero. We only need to
  % check conditions (i), (iv) and (v) of syntactic-equivalence. It is easy to
  % check (i) and (v). If either condition is false then $g$ and $h$ are not
  % syntacticly equivalent. So suppose $g$ and $h$ are two internal gates such
  % that $\Sigma (g) = \Sigma(h)$ and suppose that if they are both output gates
  % then $\Omega^{-1}(g) = \Omega^{-1}(h)$. It remains to check if $L(g)/{\sim}$
  % and $L(h)/{\sim}$ are isomorphism equivalent. First check if $H_h /{\sim} =
  % H_g /{\sim}$. If not, $g$ and $h$ are not syntactically equivalent. So
  % suppose $H_h /{\sim} = H_g /{\sim}$. If $g$ and $h$ are symmetric gates then
  % $H_h /{\sim} = H_g/{\sim}$ if, and only if, $L(g)/{\sim}$ and $L(h)/{\sim}$
  % are isomorphism equivalent, so let $g \sim h$. If $g$ and $h$ are not
  % symmetric gates then, since $C$ is transparent, $g$ and $h$ have unique
  % labels. We thus have that $L(g)/{\sim}$ and $L(h)/{\sim}$ are isomorphism
  % equivalent if, and only if, $L(g)^{-1}L(h)$ acts like an isomorphism. This
  % can be checked by first checking that the function induces a sorted
  % permutation on the universe of $g$ (which can be checked efficiently). This
  % is sufficient as $L(g)(L(g)^{-1}L(h)) = L(h)$.

  % It is easy to see that, given a reasonable binary encoding of a circuit, the
  % above procedure can be implemented so as to run in polynomial time, and that
  % the relation $\sim$ is the syntactic-equivalence relation on the gates of
  % $C$.
